---
title: "bioinformatics_prac1"
author: "Mark"
date: "06/05/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting started by improving my skills in R startting with arithmetic

```{r,math1}
1+2
10000^2
sqrt(64)
```

```{r,math2}
s <- sqrt(64)
s
a <- 2
b <- 5
c <- a*b
c
```

```{r,math3}
c(1,2,3)
sum(c(1,10,100))
mean(c(1,10,100))
median(c(1,10,100))
max(c(1,10,100))
min(c(1,10,100))
```

#saving objects in r and working on the objects
we use `<-` to save objects nd `=` can also play same role 
```{r, math4}
a <- c(1,10,100)
a
2*a
a+1
sum(a)
mean(a)
median(a)
sd(a)
var(a)
length(a)
```

Understanding R objects types
```{r, math5}
A <- c(1,10,100)
A
str(A)
class(A)
typeof(A)

B <- 1:10
B
str(B)
class(B)
typeof(B)

C <- c("prime1"=2, "prime2"=3, "prime3"=5, "prime4"=7)
C
str(C)
class(C)
typeof(C)

D <- "x1" 
D
str(D)
class(D)
typeof(D)

E <- c("x1", "y2", "z3")
E
str(E)
names(E) <- c("code1","code2","code3")
E
str(E)
class(E)
typeof(E)
```

Understanding factors which are categorically used in R for biological sex data
```{r, factors1}
x <- factor(c("single", "married", "married", "single","defacto","widowed"))
x
str(x)
levels(x)
```

```{r,factors2}
y <- c("good", "very good", "fair", "poor","good")
str(y)
yy <- factor(y, levels = c("very poor","poor","fair","good","very good"),ordered = TRUE)
yy
str(yy)
levels(yy)
```

Understanding the logical values in R which makes use of `TRUE/FALSE` a lot
```{r, logicals}
myvariable1 <- 0
as.logical(myvariable1)
myvariable2 <- 1
as.logical(myvariable2)
vals <- c(0,1,0,1,0,0,0,1,1,0,1,0,1,0)
vals
str(vals) 
as.logical(vals)
```

Understanding how to subset vectors and save the results to a new variable
```{r,vec_subset}
a <- c(2:22,98,124,3002)
a
length(a)
a[2]
a[3:4]
a[c(1,3)] 
a[length(a)]
a[(length(a)-1)]
x <- a[10:(length(a)-1)] * 2
x
```





```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
