---
title: "BioinfoPrac2"
author: "Mark"
date: "13/05/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section 2 of improving my skills in R, working with data tables 

```{r, datatables1}
head(freeny.x)
str(freeny.x)

head(mtcars)
str(mtcars)

#example of commands that can be used for matrices and not data frames
mean(freeny.x)
```

## Understanding the commands that can be used for data frames and matrices 
Find out the number of rows, columns, and simple operations on those rows and columns.
```{r,datatables_commands}

# number of columns
ncol(mtcars)

# number of rows
nrow(mtcars)

# dimensions
dim(mtcars)

# analysing rows and columns
colMeans(freeny.x)
colSums(freeny.x)
rowMeans(freeny.x)
rowSums(freeny.x)

```

## Understanding how to transpose a matrix or data frame
```{r,transpose1}

freeny_flip <- t(freeny.x)
head(freeny_flip)

mtcars_flip <- t(mtcars)
head(mtcars_flip)
str(mtcars_flip)

mtcars_flip <- as.data.frame(mtcars_flip)
str(mtcars_flip)

```

## Subsetting a data frame

Understanding how to perform filtering in data analysis
the syntax is `df[rows,cols]`

```{r,datatables2}

# get rows 1-10 of column 2
freeny.x[1:10,2]

# get rows 1-6 of columns 1-3
freeny.x[1:6,1:3]

# get rows 1-6 of all columns
freeny.x[1:6,]

# get all rows for columns 1 and 2 
freeny.x[,1:2]

```

Understanding how to subset just one column or row

```{r,datatables3}

# get all rows of column 1
freeny.x[,1]

# to prevent conversion to vector, use drop=FALSE
freeny.x[,1,drop=FALSE]

# the same concept for rows
freeny.x[1,]

# the same concept for rows
freeny.x[1,,drop=FALSE]

```

```{r,datatables4}

# get rows 1-10 of columns 1-6
mtcars[1:10,1:6]

# get all rows of column1
mtcars[,1]

# preventing conversion to vector, use drop=FALSE
mtcars[,1,drop=FALSE]

```

Subsetting data frames based on the name of the column or row using square brackets

```{r,datatables5}

mtcars[,"cyl"]

mtcars[,c("mpg","wt")]

mtcars["Camaro Z28",c("mpg","wt")]

```

Subsetting data frames columns using the `$` notation
The syntax is `df$col`.
```{r,datatables6}
mtcars$mpg
```

using the `$` notation to create new columns and rounding to 3 significant figures

```{r,datatables7}
mtcars$lper100km <- 235.215 / mtcars$mpg
mtcars[,c(1,ncol(mtcars))]
mtcars$lper100km <- signif(235.215 / mtcars$mpg ,3)
mtcars[,c(1,ncol(mtcars))]
```

subset command `subset()` is also perfect for filtering based on more than one criteria using the `&` and `|` operators.
```{r, subsetting2}
mtcars
subset(mtcars,lper100km < 10)

# you want an economical AND quick car
subset(mtcars,lper100km < 10 & qsec < 18)

# you want an economical OR quick car
subset(mtcars,lper100km < 10 | qsec < 18)
```

using the `colnames` and `rownames` to get the row or column names and modifying them.

```{r,rowcolnames1}
colnames(mtcars)
rownames(mtcars)
colnames(mtcars) <- c("miles per gallon",
              "number of cylinders",
              "displacement in cubic inches",
              "gross horsepower",
              "rear axle ratio",
              "weight (pounds/1000)",
              "quarter mile time in seconds",
              "V or straight cylinder configuration",
              "transmission type: auto (0) or manual (1)",
              "number of forward gears",
              "number of carburetors",
              "litres per 100km")
head(mtcars)
colnames(mtcars)[1] <- "miles per US gallon"
head(mtcars)
```

If having a whitespace in the column or row name, which might cause problems later on with subsetting. 
much better to wrap the column name in backticks like this.

```{r, subsetting4}
economical_cars <- subset(mtcars,`litres per 100km` < 10)
economical_cars
```

understanding how to be able to subset a data frame based on the row names.
Let's get all the Mercedes models.
To do this, we need to introduce the `grep()` command which matches strings.

```{r,rowcolnames2}
# let's look again at the car names
rownames(mtcars)

# lets filter all the ones with "Merc in the name"
grep("Merc",rownames(mtcars))

# now lets extract out all those rows
mercs <- mtcars[grep("Merc",rownames(mtcars)),]
mercs
```

## Sorting data tables

understanding how to subset economical cars by their speed based on their quarter mile time.
To do this, we need to use the `order()` command together with the square brackets.
`order()` only returns the index of the values, it doesn't actually do the sorting.
Note that `order()` default behaviour is to bring the smaller values to the top.
That can be reversed by putting a `-` before the vector being ordered.

```{r,sorting1}

economical_cars

order(economical_cars$`quarter mile time in seconds`)

sorted <- economical_cars[order(economical_cars$`quarter mile time in seconds`),]

sorted[,c(7,ncol(sorted))]

reverse_sorted <- economical_cars[order(-economical_cars$`quarter mile time in seconds`),]

reverse_sorted[,c(7,ncol(reverse_sorted))]
```

## Creating data frames and matrices

First we will create a data frame for some people who completed a survey about their height and weight.
You should always run `str()` to check that the resulting dataframe has the intended structure.
You may need to include `stringsAsFactors=FALSE` to protect character strings being converted to factors.

```{r,create_df1}
pnames <- c("Jill", "Matt", "Sam", "Amy", "Bob", "Raj")
pnames
pgender <- as.factor(c("F", "M", "F", "F", "M", "M"))
pgender
pheight <- c(164, 186, 170, 175, 178, 191)
pheight
pweight <- c(54.1, 90.3, 64.8, 66.7, 80.4, 86.9)
pweight
df <- data.frame(pnames,pgender,pheight,pweight)
str(df)
df <- data.frame(pnames,pgender,pheight,pweight,stringsAsFactors = FALSE)
str(df)
df
```

Now we might want to make the row names the name of the person.
This makes the data tidier, but it won't work if there are more then one entry with the same name.
You can use the `NULL` to delete columns.
Deleting rows can be done with square brackets.

```{r,create_df2}
rownames(df) <- df$pnames
df
df$pnames=NULL
df
# delete row 2 and 4
df <- df[-c(2,4),]
df
```

Now we will convert df into a matrix.

```{r,create_mat1}
as.matrix(df)
df$pgender <- as.numeric(df$pgender)
mymat <- as.matrix(df)
mymat
str(mymat)
```

We can also convert other types of data into a matrix.
To demonstrate this, I'll create some random data with `rnorm` and convert it into a matrix.

```{r,create_mat2}
mydata <- rnorm(n = 100, mean = 10, sd = 20)
mymatrix <- matrix(data = mydata, nrow = 20, ncol = 5)
mymatrix
```

## Creating charts from data frame and matrix objects

Making charts from data frames and matrix objects is really similar to prac1 bioinfo skills.
Here are some examples.

```{r,plots1,fig.height=8,fig.width=8}

# histogram of cylinders
data(mtcars)
hist(mtcars$cyl,xlab="number of cylinders", main="number of cylinders in mtcars data")

# boxplot of qsec values for Toyota and Mercedes
toyota <- mtcars[grep("Toyota",rownames(mtcars)),]
merc <- mtcars[grep("Merc",rownames(mtcars)),]
boxplot(toyota$qsec, merc$qsec,
        ylab="quarter mile time in seconds",
        main="mtcars",
        names = c("Toyota","Mercedes"))

# scatterplot of petal length vs sepal length for setosa irises
# include a trend line using the lm function
setosa <- subset(iris,Species=="setosa")
head(setosa)
mylm <- lm(setosa$Sepal.Length ~ setosa$Petal.Length)
plot(setosa$Petal.Length,setosa$Sepal.Length, 
     xlab="Petal length (cm)",
     ylab="Sepal length (cm)",
     main="Setosa petal and sepal length",
     pch=19)
abline(mylm,col="red",lwd=2,lty=2)

# a pairs plot is a special type of scatterplot
pairs(mymatrix)

# lets make a line diagram of freeny revenues
# first need to normalise each column to the initial value
head(freeny.x)
freeny_norm <-t( t(freeny.x) / freeny.x[1,] )
head(freeny_norm)
plot(freeny_norm[,1], type="b", ylim=c(0.9,1.12), col="blue",
     xlab="Quarters beginning 1962 Q2",
     ylab="Change in values overtime")
  lines(freeny_norm[,2], type="b", col="red")
  lines(freeny_norm[,3], type="b", col="black")
  lines(freeny_norm[,4], type="b", col="darkgreen")
  legend("topleft", legend = colnames(freeny.x), lty=1 , col = c("blue","red", "black", "darkgreen"))
  
```


In R, sometimes we need to load a particular package in order to make a special type of chart.
In the example below we are making a heatmap, where the colour indicates the numerical value.

```{r,heatmaps,fig.height=8,fig.width=8  }
library("gplots")
heatmap.2(mymatrix,trace="none",scale="none",main="heatmap")
heatmap.2(cor(mymatrix),trace="none",scale="none", main="correlation")
```

This is also a good time to let you know that you can save R charts as files.
There are different types of file formats, but PNG and PDF are the most used types.
You will see these new files appear in your files menu.

```{r,output_plots}
pdf("myplot.pdf")
plot(1:10)
dev.off()
png("myplot.png")
plot(1:10)
dev.off()
```

## Checking my understanding 

solving the questions below.

1. Create a scatterplot of mtcars weight (x axis) versus mpg (y axis). 
Include x and y axis labels and a main heading.

2. Sort mtcars by weight (wt) and create a horizontal barplot of wt values so heaviest ones at shown at the top of the bar plot.
The plot should be labeled so it is clear to see which bar belong to which car.
Include an axis label and main title.

3. Create a box plot of iris petal lengths.
Each species should be a different category.
Chart needs axis labels and a main title.

```{r, question1}
mtcars[,c("wt","mpg")]
mylm <- lm(mtcars$wt ~ mtcars$mpg)
plot(mtcars$wt,mtcars$mpg, 
     xlab="weight",
     ylab="mpg",
     main="Mtcars weight and mpg",
     pch=19)
```

```{r, question2}
mycars <- mtcars[order(mtcars$wt),]
par(mar=c(5,12,5,3))
barplot(mycars$wt, xlim= c(0,6), horiz=TRUE,las=1,names.arg=rownames(mycars),cex.names=0.6,main="mtcars weights",xlab="car weight (lb/1000)")
grid()

```


```{r, question3}
setosa <- subset(iris,Species=="setosa")
head(setosa)

virginica <- subset(iris,Species=="virginica")
head(virginica)

versicolor <- subset(iris,Species=="versicolor")
head(versicolor)

boxplot(list(setosa$Petal.Length,virginica$Petal.Length,versicolor$Petal.Length),names=c("Setosa","Virginica","Versicolor"),ylab="cm",main="Petal length")

```

## Reading in data files
understanding how to read in data files that could be in different formats like text(.txt), comma separated values (.csv), tab separated values (.tsv) and perhaps excel files too (.xls and xlsx)
using the `read.table()` command to understand some really important options.

```{r,tsv}

URL="https://raw.githubusercontent.com/markziemann/SLE712_files/master/pipette_test.tsv"
# look closely at the structure of pip
# can you see what is wrong?
pip <- read.table(URL)
pip
str(pip)
# try again
pip <- read.table(URL,stringsAsFactors = FALSE)
pip
str(pip)
# try again
# looking better
pip <- read.table(URL,stringsAsFactors = FALSE, header=TRUE)
pip
str(pip)
# got it now
pip <- read.table(URL,stringsAsFactors = FALSE, header=TRUE, row.names=1)
pip
str(pip)

```

Now let's try a csv file containing some travel records.

```{r,readcsv}
URL="https://people.sc.fsu.edu/~jburkardt/data/csv/airtravel.csv"
trav <- read.table(URL,sep=",")
trav
str(trav)
trav <- read.table(URL,sep=",",header=TRUE)
trav
str(trav)
trav <- read.csv(URL)
trav
str(trav)
```

Now let's try a Microsoft Excel file.

```{r,readxl}
URL="https://github.com/file-examples-com/uploads/blob/master/2017/02/file_example_XLS_10.xls?raw=true"
NAME="file_example_XLS_10.xls"
download.file(URL,destfile=NAME)
library("readxl")
mydata <- read_xls(NAME)
mydata
str(mydata)
```

## Save and load session and single datasets

When working in R, it is convenient to save the session with `save.image()`. 
This results in an Rdata file which contains all the data objects in your current environment.
To test that it's actually working, clear your environment with the sweep/broom icon and then load the Rdata file with `load()`.

```{r,save1}
save.image("mysession.Rdata")
rm(list=ls())
load("mysession.Rdata")
```

That is really cool, but sometimes we want to save individual objects, such as a large dataframe as Rdata files.

```{r,save2}
saveRDS(object = mymatrix , file = "mymatrix.Rds")
rm(list=ls())
x <- readRDS("mymatrix.Rds")
head(x)
```

## Check your skills

For the TSV file located here: https://raw.githubusercontent.com/markziemann/SLE712_files/master/bioinfo_asst3_part1_files/mydata.tsv do the following:

1. Read it in and show the first 6 rows of data. 

2. Calculate the column and row means.

3. Use the `cor()` command to find the correlation coefficients between the 3 data sets. Which two datasets are the most similar?

4. Make a pairs plot of the three datasets.

## Answers to TSV file

```{r, tsv1}
URL="https://raw.githubusercontent.com/markziemann/SLE712_files/master/bioinfo_asst3_part1_files/mydata.tsv"
x <- read.table(URL,header=TRUE,row.names = 1)
head(x)
```

```{r,tsv2}
colMeans(x)
rowMeans(x)
```

```{r,tsv3}
cor(x)

library("gplots")
heatmap.2(cor(x),trace="none",scale="none",margins = c(15,15))
```

```{r,tsv4}
pairs(x)
```



## Session information

For reproducibility.

```{r,sessioninfo}
sessionInfo()
```





