---
title: "bioinformatics_prac1"
author: "Mark"
date: "06/05/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting started by improving my skills in R startting with arithmetic

```{r,math1}
1+2
10000^2
sqrt(64)
```

```{r,math2}
s <- sqrt(64)
s
a <- 2
b <- 5
c <- a*b
c
```

```{r,math3}
c(1,2,3)
sum(c(1,10,100))
mean(c(1,10,100))
median(c(1,10,100))
max(c(1,10,100))
min(c(1,10,100))
```

#saving objects in r and working on the objects
we use `<-` to save objects nd `=` can also play same role 
```{r, math4}
a <- c(1,10,100)
a
2*a
a+1
sum(a)
mean(a)
median(a)
sd(a)
var(a)
length(a)
```

Understanding R objects types
```{r, math5}
A <- c(1,10,100)
A
str(A)
class(A)
typeof(A)

B <- 1:10
B
str(B)
class(B)
typeof(B)

C <- c("prime1"=2, "prime2"=3, "prime3"=5, "prime4"=7)
C
str(C)
class(C)
typeof(C)

D <- "x1" 
D
str(D)
class(D)
typeof(D)

E <- c("x1", "y2", "z3")
E
str(E)
names(E) <- c("code1","code2","code3")
E
str(E)
class(E)
typeof(E)
```

Understanding factors which are categorically used in R for biological sex data
```{r, factors1}
x <- factor(c("single", "married", "married", "single","defacto","widowed"))
x
str(x)
levels(x)
```

```{r,factors2}
y <- c("good", "very good", "fair", "poor","good")
str(y)
yy <- factor(y, levels = c("very poor","poor","fair","good","very good"),ordered = TRUE)
yy
str(yy)
levels(yy)
```

Understanding the logical values in R which makes use of `TRUE/FALSE` a lot
```{r, logicals}
myvariable1 <- 0
as.logical(myvariable1)
myvariable2 <- 1
as.logical(myvariable2)
vals <- c(0,1,0,1,0,0,0,1,1,0,1,0,1,0)
vals
str(vals) 
as.logical(vals)
```

Understanding how to subset vectors and save the results to a new variable
```{r,vec_subset}
a <- c(2:22,98,124,3002)
a
length(a)
a[2]
a[3:4]
a[c(1,3)] 
a[length(a)]
a[(length(a)-1)]
x <- a[10:(length(a)-1)] * 2
x
```

Coersing R data into different types such as conversions of objects
```{r,conv1}
a <- c(1.9,2.7,3.3,5.1,9.9,0)
a
as.integer(a)
as.character(a)
as.logical(a)
as.factor(a)

b <- c("abc","def","ghi","jkl")
b
as.numeric(b)
as.logical(b)
as.integer(b)
my_factor <- as.factor(b)
as.numeric(my_factor)
```

Creating random and semi random datas which are mostly used in simulation analysis
```{r,sample1}
nums <- 1:5
sample(x = nums, size = 3)
sample(x = nums, size = 5)
sample(x = nums ,size = 10, replace = TRUE)
```

understanding how how to sample from distributions. Example: sampling 5 numbers from a normal distribution around a median of 10 and standard deviation of 2
```{r,distributions1}
d <- rnorm(n = 5, mean = 10, sd = 2)
d
```

Example on sampling 20 numbers from a binomial distribution with a size of 50 and probability of 0.5
```{r,distributions2}
b <- rbinom(n = 20, size = 50, prob = 0.5)
b
mean(b)
```

## Including Plots

How to do a dot and line plot and also being bale to add extra lines, changing line colour and adding subheading

```{r,dot1}
a <- (1:10)^2
a
plot(a)
plot(a,type="l")
plot(a,type="b")

plot(a,type="b")
lines(a/2, type="b",col="red")
mtext("Black:Growth of A. Red: growth of A/2")
```

how to plot a scatterplots and change the point and type (`pch`) size (`cex`), as well as add a main heading (`main`). including the possibility of being able to add series of points to the chart and adjust the axis limits with `xlim` and `ylim`

```{r,scatter1}
x_vals <- rnorm(n = 1000, mean = 10, sd = 2)
x_vals
d_error <- rnorm(n = 1000, mean = 1, sd = 0.1)
d_error
y_vals <- x_vals * d_error
y_vals
plot(x=x_vals, y=y_vals, xlab="my x values", ylab="my y values")

plot(x=x_vals, y=y_vals, xlab="my x values", ylab="my y values",pch=19, cex=0.5, main="Plot of X and Y values")

plot(x=x_vals, y=y_vals, xlab="my x values", ylab="my y values",pch=19, cex=0.5, main="Plot of X and Y values")
points(x=x_vals, y=y_vals/2, pch=19, cex=0.5,col="blue")

plot(x=x_vals, y=y_vals, xlab="my x values", ylab="my y values",pch=19, 
     cex=0.5, main="Plot of X and Y values",
     ylim=c(0,17))
points(x=x_vals, y=y_vals/2, pch=19, cex=0.5,col="blue")
```

understanding how to run a linear regression

```{r,scatter2}
linear_regression_model <- lm(y_vals ~ x_vals)

summary(linear_regression_model)

SLOPE <- linear_regression_model$coefficients[1]
INTERCEPT <- linear_regression_model$coefficients[2]

HEADER <- paste("Slope:",signif(SLOPE,4),"Intercept:",signif(INTERCEPT,4))

plot(x=x_vals, y=y_vals, xlab="my x values", ylab="my y values",pch=19, cex=0.5)
abline(linear_regression_model,col="red",lty=2,lwd=3)
mtext(HEADER)
```

Plotting of boxplots

```{r,boxplots1}
boxplot(x_vals, y_vals/2)
boxplot(x_vals, y_vals/2, names=c("X values", "Y values"),ylab="Measurement (cm)")
```

plotting of histogramms and possibility of placing multiple charts on a  single image.

```{r,histograms1}
hist(x_vals)

par(mfrow = c(2, 1))
hist(x_vals,main="")
hist(y_vals,main="")
```

## Session information

For reproducibility.

```{r,sessioninfo}
sessionInfo()
```





